---
title: "Learning States Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning States Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Saqrlab)
```

## Overview

Saqrlab includes over 180 learning action verbs organized into 8 categories, designed for educational research simulations. This reference documents all available learning states and how to use them.

## Available Categories

```{r list-categories}
list_learning_categories()
```

## Complete Learning States Reference

### Metacognitive (20 verbs)

Self-regulation and awareness actions for planning, monitoring, and evaluating one's own learning.

```{r metacognitive}
get_learning_states("metacognitive")
```

### Cognitive (30 verbs)

Mental processing actions for understanding, analyzing, and synthesizing information.

```{r cognitive}
get_learning_states("cognitive")
```

### Behavioral (30 verbs)

Observable study actions including reading, writing, practicing, and note-taking.

```{r behavioral}
get_learning_states("behavioral")
```

### Social (30 verbs)

Interaction and collaboration actions for discussing, helping, teaching, and asking.

```{r social}
get_learning_states("social")
```

### Motivational (30 verbs)

Effort and persistence actions for focusing, persisting, engaging, and striving.

```{r motivational}
get_learning_states("motivational")
```

### Affective (30 verbs)

Emotional and attitudinal states including enjoying, coping, managing stress, and curiosity.

```{r affective}
get_learning_states("affective")
```

### Group Regulation (9 verbs)

Socially shared regulation of learning (SSRL) actions for collaborative planning, monitoring, and adapting.

```{r group-regulation}
get_learning_states("group_regulation")
```

### LMS Actions (30 verbs)

Learning Management System interactions including viewing content, accessing resources, and submitting assignments.

```{r lms}
get_learning_states("lms")
```

## Using Learning States

### Basic Retrieval

```{r basic-retrieval}
# Get all verbs from a category
metacognitive_verbs <- get_learning_states("metacognitive")
length(metacognitive_verbs)

# Get verbs from multiple categories
srl_verbs <- get_learning_states(c("metacognitive", "cognitive", "motivational"))
length(srl_verbs)

# Get all verbs
all_verbs <- get_learning_states("all")
length(all_verbs)
```

### Random Sampling

```{r random-sampling}
# Random 8 verbs from any category
random_8 <- get_learning_states(n = 8, seed = 42)
random_8

# Random 6 verbs from specific categories
random_cog <- get_learning_states(c("cognitive", "behavioral"), n = 6, seed = 42)
random_cog
```

### Smart Selection

The `smart_select_states()` function intelligently selects states based on network size:

```{r smart-select}
# Small network: auto-selects from one category
small_net <- smart_select_states(5, seed = 42)
small_net

# Medium network: balanced across categories
medium_net <- smart_select_states(10, seed = 42)
medium_net

# Large network: wide variety
large_net <- smart_select_states(20, seed = 42)
large_net
```

### Biased Selection

```{r biased-select}
# Prioritize metacognitive verbs
meta_focus <- smart_select_states(
  n_states = 10,
  primary_categories = "metacognitive",
  secondary_categories = "cognitive",
  primary_ratio = 0.7,  # 70% metacognitive
  seed = 42
)
meta_focus
```

## Integration with Simulation Functions

### With simulate_sequences()

```{r simulate-with-states}
# Auto-generates with learning states by default
sequences <- simulate_sequences(
  n_sequences = 50,
  seq_length = 15,
  n_states = 5,
  seed = 42
)

# Check states used
unique(unlist(sequences))
```

### Specifying Categories

```{r specify-categories}
# Use specific categories
sequences <- simulate_sequences(
  n_sequences = 50,
  seq_length = 15,
  n_states = 6,
  categories = c("metacognitive", "cognitive"),
  seed = 42
)

unique(unlist(sequences))
```

### With simulate_matrix()

```{r matrix-with-states}
# Matrix uses random category by default
mat <- simulate_matrix(n_nodes = 5, seed = 42)
rownames(mat)

# Run again - different category selected
mat2 <- simulate_matrix(n_nodes = 5, seed = 123)
rownames(mat2)
```

### With simulate_htna()

```{r htna-with-states}
# HTNA uses different category per type
net <- simulate_htna(seed = 42)
lapply(net$node_types, head, 3)
```

## Direct Access to LEARNING_STATES

```{r direct-access}
# Access the raw data
names(LEARNING_STATES)

# Get specific category
LEARNING_STATES$metacognitive

# Count verbs per category
sapply(LEARNING_STATES, length)
```

## Global Names Dataset

Saqrlab also includes 300 diverse names for simulation:

```{r global-names}
# Access names
head(GLOBAL_NAMES, 20)

# Get random names
get_global_names(n = 10, seed = 42)
```

## Educational Research Applications

### Self-Regulated Learning (SRL)

```{r srl-application}
# Zimmerman's SRL phases
forethought <- get_learning_states("metacognitive")
performance <- get_learning_states(c("cognitive", "behavioral"))
reflection <- c("Evaluate", "Reflect", "Assess", "Judge")

cat("Forethought:", head(forethought, 5), "...\n")
cat("Performance:", head(performance, 5), "...\n")
cat("Reflection:", reflection, "\n")
```

### Socially Shared Regulation of Learning (SSRL)

```{r ssrl-application}
# SSRL components
ssrl_verbs <- get_learning_states("group_regulation")
ssrl_verbs
```

### Learning Analytics (LMS Data)

```{r la-application}
# Common LMS actions
lms_verbs <- get_learning_states("lms")
head(lms_verbs, 15)
```

### Collaborative Learning

```{r collab-application}
# Social and group regulation
collab_verbs <- get_learning_states(c("social", "group_regulation"))
length(collab_verbs)
head(collab_verbs, 10)
```

## Custom State Names

You can always use your own state names:

```{r custom-names}
# With simulate_sequences
sequences <- simulate_sequences(
  n_sequences = 50,
  seq_length = 15,
  n_states = 4,
  states = c("Explore", "Learn", "Practice", "Master"),
  seed = 42
)

unique(unlist(sequences))

# With simulate_matrix
mat <- simulate_matrix(
  n_nodes = 4,
  names = c("Phase1", "Phase2", "Phase3", "Phase4"),
  seed = 42
)

rownames(mat)
```

## Complete Reference Table

```{r reference-table}
# Create comprehensive reference
ref_table <- data.frame(
  Category = names(LEARNING_STATES),
  Count = sapply(LEARNING_STATES, length),
  First_5 = sapply(LEARNING_STATES, function(x) {
    paste(head(x, 5), collapse = ", ")
  }),
  row.names = NULL
)

ref_table
```

## Summary

### Key Functions

| Function | Purpose |
|----------|---------|
| `LEARNING_STATES` | Raw dataset of all verbs |
| `get_learning_states()` | Retrieve verbs by category |
| `list_learning_categories()` | Show category summary |
| `smart_select_states()` | Intelligent selection |
| `GLOBAL_NAMES` | 300 diverse names |
| `get_global_names()` | Retrieve names |

### Category Summary

| Category | Count | Focus |
|----------|-------|-------|
| metacognitive | 20 | Self-regulation |
| cognitive | 30 | Mental processing |
| behavioral | 30 | Observable actions |
| social | 30 | Interaction |
| motivational | 30 | Effort/persistence |
| affective | 30 | Emotions |
| group_regulation | 9 | SSRL |
| lms | 30 | System interactions |
| **Total** | **209** | |

### Usage Tips

1. **Default behavior**: Most functions use learning states automatically
2. **Categories parameter**: Control which categories to sample from
3. **Custom names**: Always supported via `states` or `names` parameter
4. **Reproducibility**: Use `seed` parameter for consistent results
5. **HTNA/MLNA**: Each type automatically gets different category
