---
title: "Saqrlab Reference — Missing-data mechanisms"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers `inject_missingness()`, a *transformer* that takes a
complete data frame and injects `NA`s under a chosen missingness mechanism from
the Rubin taxonomy — MCAR (completely at random), MAR (at random, driven by an
observed predictor), and MNAR (not at random, driven by the missing value
itself). The returned object carries a `"missing_info"` attribute recording
exactly which cells were made missing.

## `inject_missingness()`

Inject missing values into target columns under a known mechanism. MAR requires
an observed `predictor` column (which is itself never made missing); MNAR sets
larger numeric values missing more often. All three mechanisms are calibrated so
the realized missing fraction is approximately `prop`.

**Signature**
```{r}
args(inject_missingness)
```

We start from a small complete data frame with a regression structure so that
`x1` can serve as a MAR predictor of `y`.

```{r}
df <- simulate_prediction(
  n = 500,
  coefs = c("(Intercept)" = 1, x1 = 2, x2 = -1),
  seed = 1
)$data
head(df)
colSums(is.na(df))   # no missingness yet
```

**MCAR — missing completely at random**

Every cell of the target columns is independently set missing with probability
`prop`.
```{r}
mcar <- inject_missingness(df, mechanism = "MCAR", prop = 0.2, seed = 7)
colSums(is.na(mcar))
attr(mcar, "missing_info")$realized_prop
str(attr(mcar, "missing_info"), max.level = 1)
```

**MAR — missing at random (driven by observed `x1`)**

Here only `y` is targeted, and its missingness probability depends on the
*observed* predictor `x1` — never on `y`'s own value.
```{r}
mar <- inject_missingness(df, mechanism = "MAR", prop = 0.2,
                          cols = "y", predictor = "x1", seed = 7)
colSums(is.na(mar))
attr(mar, "missing_info")$realized_prop
```

The dependence on the predictor is visible: rows with missing `y` have a
systematically different mean `x1` than rows with observed `y`.
```{r}
miss_y <- is.na(mar$y)
tapply(df$x1, ifelse(miss_y, "y missing", "y observed"), mean)
```

**MNAR — missing not at random (driven by `y` itself)**

The probability that `y` is missing depends on `y`'s own (latent) value: larger
values are more likely to go missing.
```{r}
mnar <- inject_missingness(df, mechanism = "MNAR", prop = 0.2,
                           cols = "y", seed = 7)
colSums(is.na(mnar))
attr(mnar, "missing_info")$realized_prop
```

The MNAR signature: the *observed* (non-missing) `y` mean is shifted relative to
the full-data mean, because the largest values were preferentially removed.
```{r}
c(full_mean     = mean(df$y),
  observed_mean = mean(mnar$y, na.rm = TRUE))
```

## Visualising the missingness pattern

An `image()` of the missing-cell indicator shows where the `NA`s landed. Under
MCAR they scatter uniformly across all columns; under MAR/MNAR they concentrate
in the single targeted column `y`.

```{r, fig.height=4}
op <- par(mfrow = c(1, 3), mar = c(3, 4, 3, 1))
show_pattern <- function(x, ttl) {
  m <- is.na(x[, c("y", "x1", "x2")])
  image(seq_len(ncol(m)), seq_len(nrow(m)), t(m),
        col = c("#ece7f2", "#d95f0e"), axes = FALSE,
        xlab = "", ylab = "row", main = ttl)
  axis(1, at = seq_len(ncol(m)), labels = colnames(m))
}
show_pattern(mcar, "MCAR")
show_pattern(mar,  "MAR (y ~ x1)")
show_pattern(mnar, "MNAR (y ~ y)")
par(op)
```
