---
title: "Saqrlab Reference — Random-parameter generation"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers `simulate_data()`, the seed-driven random-parameter
generator. Unlike the explicit-parameter simulators, it returns a **bare**
`data.frame` (not a `saqr_sim`), and the seed determines both the random data
*and* the structural parameters (sample size, effect sizes, number of
groups/variables) — so each seed yields a structurally unique, ready-to-analyse
dataset. It also supports edge-case ("complexity") injection and batch
generation.

## `simulate_data()`

A single dispatcher over many dataset types. Each dataset carries `type`,
`info`, and `complexity` attributes; the `info` attribute is the intended
analysis call for that type.

**Signature**
```{r}
args(simulate_data)
```

The full catalogue of types:
```{r}
.SIM_VALID_TYPES
```

### Core types

Each of these returns a bare `data.frame` ready for its intended analysis.

```{r}
d_ttest <- simulate_data("ttest", seed = 42)
str(d_ttest)
head(d_ttest)
attributes(d_ttest)[c("type", "info", "complexity")]
```

```{r}
d_anova <- simulate_data("anova", seed = 42)
str(d_anova)
table(d_anova$group)
```

```{r}
d_cor <- simulate_data("correlation", seed = 42)
str(d_cor)
round(cor(d_cor), 2)
```

```{r}
d_clusters <- simulate_data("clusters", seed = 42)
str(d_clusters)
table(d_clusters$true_cluster)
```

```{r}
d_fa <- simulate_data("factor_analysis", seed = 42)
str(d_fa)
attr(d_fa, "n_factors")
```

```{r}
d_pred <- simulate_data("prediction", seed = 42)
str(d_pred)
head(d_pred)
```

### Newer types

The dispatcher also covers `"probit"`, `"count"`, `"meta"`, `"reliability"`,
`"gam"`, `"nma"`, `"power"`, and `"cthmm"`. Their `info` attribute documents the
intended analysis. (`"power"` returns a one-row list of inputs rather than a
data frame.)

```{r}
newer <- c("probit", "count", "meta", "reliability",
           "gam", "nma", "power", "cthmm")
do.call(rbind, lapply(newer, function(tp) {
  d <- simulate_data(tp, seed = 42)
  data.frame(type = tp,
             class = class(d)[1],
             intended_analysis = attr(d, "info"),
             row.names = NULL)
}))
```

One worked example — a Poisson `count` dataset, ready for `glm(..., poisson)`:
```{r}
d_count <- simulate_data("count", seed = 42)
str(d_count)
head(d_count)
```

### Complexity (edge-case) injection

Pass `complexity` to inject specific edge cases for stress testing. With
`complexity = "na"`, missing values are introduced; with `"outliers"`, extreme
values are added. The injected cases are recorded in the `complexity` attribute.

```{r}
d_na <- simulate_data("ttest", seed = 1, complexity = "na")
sum(is.na(d_na))
attr(d_na, "complexity")
```

```{r}
d_out <- simulate_data("prediction", seed = 1, complexity = "outliers")
attr(d_out, "complexity")
summary(d_out$y)
```

You can request several at once, e.g. `complexity = c("na", "outliers")`:
```{r}
d_multi <- simulate_data("ttest", seed = 1, complexity = c("na", "outliers"))
attr(d_multi, "complexity")
```

### Batch generation

Pass `n_batch` to get a list of structurally distinct datasets (each with its
own seed) instead of a single one. In batch mode the default complexity is
`"auto"`, which randomly injects 0–3 edge cases per dataset (seed-driven, so
fully reproducible).

```{r}
batch <- simulate_data("ttest", seed = 1, n_batch = 20)
class(batch)
length(batch)
str(batch[[1]])
attr(batch[[1]], "complexity")
```

Setting `type = "batch"` generates a batch of every type at once:
```{r}
all_batches <- simulate_data("batch", seed = 1, n_batch = 5)
class(all_batches)
names(all_batches)
vapply(all_batches, length, integer(1))
```
