---
title: "Saqrlab Reference — Networks & graphs"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

These functions build social-network and graph structures with named nodes drawn
from the package's diverse name and learning-state pools. `simulate_igraph()` and
`simulate_network()` return ready-to-plot graph objects across several classic
random-graph models, while `simulate_edge_list()`, `simulate_long_data()`, and
`simulate_onehot_data()` return tidy data frames for downstream analysis.

## `simulate_igraph()`

Create an `igraph` object using a chosen random-graph model (Erdős–Rényi,
Barabási–Albert, Watts–Strogatz, SBM, and more), with named vertices and optional
edge weights.

**Signature**
```{r}
args(simulate_igraph)
```

**Example — Erdős–Rényi**
```{r, eval=requireNamespace("igraph", quietly=TRUE)}
g_er <- simulate_igraph(n = 15, model = "er", p = 0.15, seed = 42)
g_er
plot(g_er, vertex.size = 18, vertex.label.cex = 0.7, main = "Erdos-Renyi (n = 15)")
```

**Example — Barabási–Albert (scale-free)**
```{r, eval=requireNamespace("igraph", quietly=TRUE)}
g_ba <- simulate_igraph(n = 20, model = "ba", m_ba = 2, seed = 42)
plot(g_ba, vertex.size = 14, vertex.label.cex = 0.6, main = "Barabasi-Albert (n = 20)")
```

**Example — Watts–Strogatz (small-world)**
```{r, eval=requireNamespace("igraph", quietly=TRUE)}
g_ws <- simulate_igraph(n = 20, model = "ws", nei = 2, p_rewire = 0.1, seed = 42)
plot(g_ws, vertex.size = 14, vertex.label.cex = 0.6, main = "Watts-Strogatz (n = 20)")
```

## `simulate_network()`

Create a statnet `network` object (same model menu as `simulate_igraph()`) with
named vertices and optional edge weights.

**Signature**
```{r}
args(simulate_network)
```

**Example**
```{r, eval=requireNamespace("network", quietly=TRUE) && requireNamespace("sna", quietly=TRUE)}
net <- simulate_network(n = 15, model = "er", seed = 42)
net
plot(net, displaylabels = TRUE, label.cex = 0.7, main = "statnet network (n = 15)")
```

## `simulate_edge_list()`

Generate a directed/undirected weighted edge list for social-network analysis,
with each node assigned to one of several classes.

**Signature**
```{r}
args(simulate_edge_list)
```

**Example**
```{r}
edges <- simulate_edge_list(n_nodes = 12, n_classes = 3, seed = 42)
str(edges)
head(edges)
```

## `simulate_long_data()`

Generate hierarchical long-format sequence data (one row per actor-event) with
Actor, Achiever, Group, Course, Time, and Action columns — the canonical format
for educational-process mining.

**Signature**
```{r}
args(simulate_long_data)
```

**Example**
```{r}
long <- simulate_long_data(n_groups = 3, n_actors = 2, seed = 42)
str(long)
head(long)
```

## `simulate_onehot_data()`

Generate the same hierarchical sequence structure as `simulate_long_data()` but
with each state expanded into its own one-hot indicator column.

**Signature**
```{r}
args(simulate_onehot_data)
```

**Example**
```{r}
onehot <- simulate_onehot_data(n_groups = 3, n_actors = 2, seed = 42)
dim(onehot)
str(onehot, list.len = 12)
head(onehot)
```
