---
title: "Saqrlab Reference — Reference data & utilities"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

Saqrlab ships two curated reference datasets — `LEARNING_STATES` (learning-action
verbs across 8 categories) and `GLOBAL_NAMES` (culturally diverse names by region)
— plus accessor and selection helpers that turn them into ready-to-use state labels
and node names. A small validation utility, `validate_sim_params()`, standardizes
simulation parameter lists.

## `LEARNING_STATES`

A named list of learning-action verbs organized into 8 categories drawn from the
learning-sciences literature.

**Structure**
```{r}
length(LEARNING_STATES)
names(LEARNING_STATES)
lengths(LEARNING_STATES)
lapply(LEARNING_STATES, head, 5)
```

## `GROUP_REGULATION_ACTIONS`

A character vector of socially-shared regulation of learning (SSRL) action verbs,
convenient as a ready-made state set for group-regulation simulations.

**Structure**
```{r}
length(GROUP_REGULATION_ACTIONS)
GROUP_REGULATION_ACTIONS
```

## `GLOBAL_NAMES`

A named list of ~1000 culturally diverse first names, grouped by world region, for
naming actors/nodes in simulations.

**Structure**
```{r}
length(GLOBAL_NAMES)
head(names(GLOBAL_NAMES), 10)
lapply(GLOBAL_NAMES[c("western_europe", "arab", "east_asia")], head, 5)
```

## `get_learning_states()`

Retrieve learning verbs by category, optionally taking a random sample and
combining several categories.

**Signature**
```{r}
args(get_learning_states)
```

**Example**
```{r}
# All category names
names(get_learning_states())

# Cognitive verbs only
head(get_learning_states("cognitive"))

# Random 6 verbs across two categories (seeded)
get_learning_states(c("social", "cognitive"), n = 6, seed = 1)
```

## `select_states()` / `smart_select_states()`

Intelligently select a set of `n_states` labels, balancing a primary and secondary
category by `primary_ratio`. `smart_select_states()` is an alias for `select_states()`.

**Signature**
```{r}
args(select_states)
args(smart_select_states)
```

**Example**
```{r}
# Auto-strategy: number of categories scales with n_states
select_states(8, seed = 42)

# Explicit primary category
smart_select_states(5, primary_categories = "metacognitive", seed = 1)
```

## `list_learning_categories()`

Return a tidy data frame summarizing each learning-state category with its verb
count and example verbs.

**Signature**
```{r}
args(list_learning_categories)
```

**Example**
```{r}
list_learning_categories()
```

## `get_global_names()`

Retrieve diverse first names, optionally sampling `n` of them and restricting to
one or more regions.

**Signature**
```{r}
args(get_global_names)
```

**Example**
```{r}
# 8 names from the Arab region (seeded)
get_global_names(n = 8, regions = "arab", seed = 1)

# Sample across all regions
get_global_names(n = 6, seed = 1)
```

## `list_name_regions()`

Return a tidy data frame summarizing each name region with its count and example
names.

**Signature**
```{r}
args(list_name_regions)
```

**Example**
```{r}
head(list_name_regions())
```

## `validate_sim_params()`

Validate and standardize a list of simulation parameters, applying defaults and
enforcing relationships (e.g. `min_na <= max_na <= seq_length`). Both new and
legacy parameter names are accepted.

**Signature**
```{r}
args(validate_sim_params)
```

**Example — valid call**
```{r}
validate_sim_params(list(
  seq_length  = 50,
  n_sequences = 200,
  min_na      = 2,
  max_na      = 10
))
```

**Example — invalid call (wrapped in `try()` so the doc still renders)**
```{r}
try(validate_sim_params("not_a_list"))
```
