---
title: "Saqrlab Reference — Network comparison & model fitting"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers Saqrlab's model-fitting and network-comparison verbs. The
fitting helper (`fit_network_model()`) wraps the `tna` package, the pairwise
comparators (`compare_networks()`, `compare_centralities()`,
`compare_edge_recovery()`) score how similar two networks are, and the
simulation-driven comparators (`compare_estimation()`,
`compare_network_estimation()`, `compare_reliability()`, `cross_validate_tna()`)
repeatedly fit models to assess estimation stability and recovery.

Note: the pairwise `compare_*` verbs require both networks to share the same set
of states (the same alphabet), so every example below builds matched-state
inputs.

## `fit_network_model()`

Fit a TNA model (`tna`, `ftna`, `ctna`, `atna`, or `group_tna`) to wide-format
sequence data using a single unified interface over the `tna` package.

**Signature**
```{r}
args(fit_network_model)
```

**Example**
```{r}
set.seed(1)
trans_mat <- matrix(c(
  0.7, 0.2, 0.1,
  0.3, 0.5, 0.2,
  0.2, 0.3, 0.5
), nrow = 3, byrow = TRUE)
rownames(trans_mat) <- colnames(trans_mat) <- c("A", "B", "C")
init_probs <- c(A = 0.5, B = 0.3, C = 0.2)

sequences <- simulate_sequences(
  transition_matrix = trans_mat,
  initial_probabilities = init_probs,
  max_seq_length = 15,
  num_rows = 40
)

model_tna <- fit_network_model(sequences, "tna")
class(model_tna)
```

## `compare_networks()`

Compare two fitted networks (or weight-matrix lists) over a shared set of states,
returning correlation, RMSE, and edge-difference metrics plus an edge-level table.

**Signature**
```{r}
args(compare_networks)
```

**Example**
```{r}
states <- c("A", "B", "C")
mk <- function(s) {
  set.seed(s)
  list(weights = matrix(runif(9), 3, 3, dimnames = list(states, states)))
}

cmp <- compare_networks(mk(1), mk(2))
cmp$metrics
head(cmp$edge_comparison)
```

## `compare_centralities()`

Compare centrality measures (e.g. OutStrength, InStrength, Betweenness) between
two fitted `tna` models built over a shared alphabet, via correlation and rank
agreement.

**Signature**
```{r}
args(compare_centralities)
```

**Example**
```{r}
mkseq <- function(s) {
  set.seed(s)
  as.data.frame(matrix(sample(c("A", "B", "C"), 20 * 8, replace = TRUE), 20, 8))
}
t1 <- tna::tna(mkseq(1))
t2 <- tna::tna(mkseq(2))

cc <- compare_centralities(t1, t2)
cc$correlations
```

## `compare_edge_recovery()`

Score how well one network recovers the edges of another (treated as ground
truth), returning a confusion-matrix style summary: precision, recall, F1,
accuracy, and Jaccard.

**Signature**
```{r}
args(compare_edge_recovery)
```

**Example**
```{r}
states <- c("A", "B", "C")
mk <- function(s) {
  set.seed(s)
  list(weights = matrix(runif(9), 3, 3, dimnames = list(states, states)))
}

rec <- compare_edge_recovery(mk(1), mk(2))
rec$precision
rec$recall
rec$f1_score
```

## `calculate_edge_recovery()`

Alias of `compare_edge_recovery()` — identical behaviour and return value,
provided for naming convenience.

**Signature**
```{r}
args(calculate_edge_recovery)
```

**Example**
```{r}
states <- c("A", "B", "C")
mk <- function(s) {
  set.seed(s)
  list(weights = matrix(runif(9), 3, 3, dimnames = list(states, states)))
}

identical(
  calculate_edge_recovery(mk(1), mk(2)),
  compare_edge_recovery(mk(1), mk(2))
)
```

## `compare_network_estimation()`

Compare multiple TNA model types on a single dataset by repeated independent-subset
sampling, ranking them by mean Pearson correlation between sample and remaining
models.

**Signature**
```{r}
args(compare_network_estimation)
```

**Example**
```{r}
set.seed(1)
seqs <- simulate_sequences(
  transition_matrix = trans_mat,
  initial_probabilities = init_probs,
  max_seq_length = 15,
  num_rows = 40
)

ne <- compare_network_estimation(
  seqs,
  model_types = c("tna", "ftna"),
  iterations = 2,
  seed = 1,
  verbose = FALSE
)
ne$ranking
ne$winner
```

## `compare_tna_models()`

Backward-compatibility alias of `compare_network_estimation()` — same arguments
and return value.

**Signature**
```{r}
args(compare_tna_models)
```

**Example**
```{r}
identical(compare_tna_models, compare_network_estimation)
```

## `compare_estimation()`

Run many simulations comparing how well different TNA model types recover a known
ground-truth transition structure (via `tna::compare()`), returning per-model
summaries, a ranking, and the winning model.

**Signature**
```{r}
args(compare_estimation)
```

**Example**
```{r}
ce <- compare_estimation(
  models = c("tna", "ftna"),
  n_simulations = 2,
  n_sequences = 20,
  seq_length = 12,
  n_states = 3,
  seed = 1,
  verbose = FALSE,
  parallel = FALSE
)
ce$summary
ce$winner
```

## `compare_reliability()`

Assess split-half reliability (`tna::reliability()`) across simulated data
conditions sampled from user-specified ranges, returning a
`tna_reliability_comparison` object.

**Signature**
```{r}
args(compare_reliability)
```

**Example**
```{r}
cr <- compare_reliability(
  n_simulations = 2,
  n_sequences = 30,
  seq_length = 12,
  n_states = 3,
  reliability_iter = 5,
  seed = 1,
  verbose = FALSE,
  parallel = FALSE
)
cr$summary
```

## `cross_validate_tna()`

Cross-validate several TNA model types on the same data by repeated sampling,
returning per-model-type sampling results for performance comparison.

**Signature**
```{r}
args(cross_validate_tna)
```

**Example**
```{r}
cv <- cross_validate_tna(
  seqs,
  model_types = c("relative", "frequency"),
  iterations = 2,
  seed = 1,
  verbose = FALSE
)
names(cv)
```
