---
title: "Saqrlab Reference — Statistical simulators"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

The statistical simulators generate datasets with fully specified ground-truth
parameters for the most common analyses: two-group comparisons, one-way ANOVA,
correlated multivariate normal data, Gaussian-mixture clusters, and regression
prediction. Each function returns a `saqr_sim` object with `$data` (a tidy
`data.frame`) and `$params` (the true parameters used to generate it), so you
can verify that an estimator recovers the truth.

## `simulate_ttest()`

Generate data for a two-group comparison with fully specified ground-truth
means and standard deviations, designed so that `t.test(score ~ group)`
recovers the true difference at large `n`.

**Signature**
```{r}
args(simulate_ttest)
```

**Example**
```{r}
r <- simulate_ttest(n_a = 50, n_b = 50, mean_a = 100, mean_b = 105, seed = 1)
head(r$data)
t.test(score ~ group, data = r$data)
r$params
```

## `simulate_anova()`

Generate data for a one-way ANOVA with fully specified group means and standard
deviations, designed so that `aov(score ~ group)` recovers the true group
effects at large `n`.

**Signature**
```{r}
args(simulate_anova)
```

**Example**
```{r}
r <- simulate_anova(n = 30, means = c(10, 12, 15), seed = 1)
head(r$data)
summary(aov(score ~ group, data = r$data))
r$params
```

## `simulate_correlation()`

Generate multivariate normal data from an explicit correlation (or covariance)
matrix, designed so that `cor(r$data)` recovers `sigma` at large `n`.

**Signature**
```{r}
args(simulate_correlation)
```

**Example**
```{r}
R <- matrix(c(1, 0.6, 0.3,
              0.6, 1, 0.5,
              0.3, 0.5, 1), nrow = 3)
r <- simulate_correlation(n = 200, sigma = R, seed = 1)
head(r$data)
cor(r$data)            # should approximate R
r$params$sigma         # the true matrix
```

## `simulate_clusters()`

Generate multivariate data from a mixture of Gaussians with fully specified
cluster centers and standard deviations — ground truth is known, so clustering
algorithms can be benchmarked against it.

**Signature**
```{r}
args(simulate_clusters)
```

**Example**
```{r}
centers <- matrix(c(0, 0,
                    5, 5,
                    10, 0), nrow = 3, byrow = TRUE)
r <- simulate_clusters(n = 300, centers = centers, seed = 1)
head(r$data)
str(r$params)
```

```{r clusters-plot, fig.cap="First two columns coloured by true cluster."}
plot(x2 ~ x1, data = r$data, col = r$data$true_cluster, pch = 19,
     main = "simulate_clusters(): three Gaussian clusters")
```

## `simulate_prediction()`

Generate a regression dataset with continuous and categorical predictors, a
non-linear term, and known ground-truth coefficients — a richer version of
`simulate_regression()` for testing prediction workflows.

**Signature**
```{r}
args(simulate_prediction)
```

**Example**
```{r}
r <- simulate_prediction(
  n = 200,
  coefs = c("(Intercept)" = 5, x1 = 2, x2 = -1),
  cat_levels = list(treatment = c("control", "drug_a", "drug_b")),
  cat_effects = list(treatment = c(0, 3, 5)),
  error_sd = 2, seed = 42
)
head(r$data)
summary(lm(y ~ ., data = r$data))
str(r$params)
```
