---
title: "Saqrlab Reference — Item Response Theory"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers `simulate_irt()`, which generates item-response data under
a known Item Response Theory model with ground-truth item and person
parameters. It supports the 1PL (Rasch), 2PL, and 3PL dichotomous models and
the graded response model (GRM) for ordered polytomous items. Each example
below is small, fast, and seeded so the results are reproducible.

## `simulate_irt()`

Generate item-response data under a known IRT model. The returned `saqr_sim`
object carries the simulated `$data` (a person-by-item response grid) and the
true item/person parameters in `$params`, so parameter recovery can be checked
directly.

**Signature**
```{r}
args(simulate_irt)
```

**Example — 1PL (Rasch)**

In the 1PL model every item shares the same discrimination (fixed at 1); items
differ only in difficulty.
```{r}
fit_1pl <- simulate_irt(n_persons = 300, n_items = 10, model = "1PL", seed = 1)
fit_1pl
head(fit_1pl$data)
str(fit_1pl$params, max.level = 1)
```

**Example — 2PL**

The 2PL model adds an item-specific discrimination parameter `a`.
```{r}
fit_2pl <- simulate_irt(n_persons = 400, n_items = 15, model = "2PL", seed = 2)
head(fit_2pl$data)
str(fit_2pl$params, max.level = 1)
```

**Example — 3PL**

The 3PL model adds a lower-asymptote (guessing) parameter `c` (auto-set to 0.2).
```{r}
fit_3pl <- simulate_irt(n_persons = 400, n_items = 12, model = "3PL", seed = 3)
head(fit_3pl$data)
fit_3pl$params$c
```

**Example — GRM (graded response, 4 ordered categories)**

For polytomous items, responses take values `0 ... n_categories - 1` and the
true category thresholds are returned in `$params$thresholds`.
```{r}
fit_grm <- simulate_irt(n_persons = 400, n_items = 8, model = "GRM",
                        n_categories = 4, seed = 4)
head(fit_grm$data)
str(fit_grm$params, max.level = 1)
fit_grm$params$thresholds
```

## Parameter recovery

For a 2PL fit, item proportion-correct should track true item difficulty `b`
negatively (easier items are answered correctly more often), and person total
score should track true ability `theta` positively.

```{r}
prop_correct <- colMeans(as.data.frame(fit_2pl))
cor(prop_correct, fit_2pl$params$b)        # strongly negative

total_score <- rowSums(as.data.frame(fit_2pl))
cor(total_score, fit_2pl$params$theta)     # strongly positive
```

A plot of item proportion-correct against true difficulty `b` makes the
recovery visible: the points fall along a clear downward trend.

```{r}
plot(fit_2pl$params$b, prop_correct,
     pch = 19, col = "#2c7fb8",
     xlab = "True item difficulty  b",
     ylab = "Observed proportion correct",
     main = sprintf("2PL recovery: cor = %.2f",
                    cor(prop_correct, fit_2pl$params$b)))
abline(lm(prop_correct ~ fit_2pl$params$b), col = "#d95f0e", lwd = 2)
```

## Optional: estimating with `mirt`

When the `mirt` package is installed, the simulated data can be fit to confirm
the generating parameters are recoverable by an external estimator. This chunk
is skipped if `mirt` is not available.

```{r mirt-recovery, eval=requireNamespace("mirt", quietly=TRUE)}
mod <- mirt::mirt(as.data.frame(fit_2pl), 1, itemtype = "2PL", verbose = FALSE)
est_b <- mirt::coef(mod, simplify = TRUE, IRTpars = TRUE)$items[, "b"]
cor(est_b, fit_2pl$params$b)     # estimated vs. true difficulty
```
