---
title: "Saqrlab Reference — TNA simulation"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

The TNA-simulation family generates synthetic Temporal Network Analysis data with known ground-truth parameters. The `simulate_tna_*` functions build fitted `tna`/`group_tna` model objects (ready for the `tna` package), `simulate_*_matrix`/`simulate_htna` return transition matrices with node groupings for hierarchical/multilevel plots, and `generate_probabilities` produces the underlying Markov probabilities. The `generate_*` names are silent backward-compatibility aliases of their `simulate_*` counterparts and produce identical output. All examples below are seeded and kept small for fast rendering.

## `simulate_tna_network()`

Generate a single fitted TNA network model with learning-state names — the simplest way to obtain a ready-to-use `tna` object (`$weights`, `$inits`, `$labels`, `$data`).

**Signature**
```{r}
args(simulate_tna_network)
```

**Example**
```{r}
model <- simulate_tna_network(n_states = 4, seed = 1)
class(model)
names(model)
head(round(model$weights, 3))     # transition weights; rows sum to 1
round(rowSums(model$weights), 3)
# tna ships a plot method; guard it in case the suggested deps are absent
invisible(try(plot(model), silent = TRUE))
```

## `simulate_tna_networks()`

Simulate multiple independent TNA network objects for simulation studies, returning a named list of fitted `tna` models (no underlying data/probabilities). Alias: `generate_tna_networks()`.

**Signature**
```{r}
args(simulate_tna_networks)
```

**Example**
```{r}
nets <- simulate_tna_networks(
  n_networks = 2, n_states = 4, n_sequences = 30, seq_length = 10,
  seed = 1, verbose = FALSE
)
length(nets)
names(nets)
class(nets[[1]])
head(round(nets[[1]]$weights, 3))
```

## `simulate_tna_datasets()`

Simulate complete TNA datasets: each element bundles the fitted `model`, the generating `transition_probs`/`initial_probs`, the simulated `sequences`, and a `params` list. Aliases: `generate_tna_datasets()`, `generate_sequence_data()`.

**Signature**
```{r}
args(simulate_tna_datasets)
```

**Example**
```{r}
data <- simulate_tna_datasets(
  n_datasets = 2, n_states = 4, n_sequences = 30, seq_length = 10,
  seed = 1, verbose = FALSE
)
length(data)
names(data)
names(data[[1]])                       # model, params, probs, sequences
data[[1]]$params$state_names
head(data[[1]]$sequences, 3)
round(rowSums(data[[1]]$transition_probs), 3)
```

## `simulate_group_tna_networks()`

Simulate a grouped/clustered TNA model (class `group_tna`) — one TNA network per group (e.g. classrooms or teams), built from long-format sequence data. Alias: `generate_group_tna_networks()`.

**Signature**
```{r}
args(simulate_group_tna_networks)
```

**Example**
```{r}
group_net <- simulate_group_tna_networks(
  n_groups = 2, n_actors = 5, n_states = 3, seq_length_range = c(5, 8),
  seed = 1, verbose = FALSE
)
class(group_net)
length(group_net)                      # one tna model per group
names(group_net)
head(round(group_net[[1]]$weights, 3))
```

## `simulate_tna_matrix()`

Simulate a transition matrix with node groupings, compatible with `tna::plot_htna()` (hierarchical) and `tna::plot_mlna()` (multilevel). Returns `list(matrix, node_types)`. A convenience wrapper around `simulate_htna()`. Alias: `generate_tna_matrix()`.

**Signature**
```{r}
args(simulate_tna_matrix)
```

**Example**
```{r}
net <- simulate_tna_matrix(
  nodes_per_group = 2, n_groups = 2,
  group_names = c("Macro", "Micro"),
  seed = 1, verbose = FALSE
)
names(net)
dim(net$matrix)
round(net$matrix, 3)
round(rowSums(net$matrix), 3)          # transition matrix: rows sum to 1
net$node_types                         # group -> node-name mapping
```

## `simulate_htna()`

Simulate a transition matrix with multiple node *types* for hierarchical (HTNA), multilevel (MLNA), or multi-type (MTNA) analysis. Returns `list(matrix, node_types, type_names, n_nodes_per_type)`. `simulate_mlna()` and `simulate_mtna()` are aliases that produce identical output — choose the name matching your analysis context.

**Signature**
```{r}
args(simulate_htna)
```

**Example**
```{r}
htna <- simulate_htna(n_nodes = 2, n_types = 2, seed = 1)
names(htna)
htna$type_names
dim(htna$matrix)
round(htna$matrix, 3)
round(rowSums(htna$matrix), 3)
# a quick heatmap of the transition structure
image(htna$matrix, axes = FALSE, main = "simulate_htna() transition matrix")
```

## `simulate_mlna()`

Multilevel-network alias of `simulate_htna()` — identical behaviour and return value; the name simply documents intent in multilevel (MLNA) workflows.

**Signature**
```{r}
args(simulate_mlna)
```

**Example**
```{r}
mlna <- simulate_mlna(
  n_nodes = 2,
  type_names = c("Macro", "Meso", "Micro"),
  seed = 1
)
mlna$type_names
dim(mlna$matrix)
round(rowSums(mlna$matrix), 3)
```

## `simulate_mtna()`

Multi-type-network alias of `simulate_htna()` — identical behaviour and return value, named for multi-type (MTNA) workflows.

**Signature**
```{r}
args(simulate_mtna)
```

**Example**
```{r}
mtna <- simulate_mtna(n_nodes = 2, n_types = 3, seed = 1)
mtna$type_names
dim(mtna$matrix)
round(rowSums(mtna$matrix), 3)
```

## `simulate_matrix()`

Generate a single simple network matrix (transition, frequency, co-occurrence, or adjacency) with learning-state node names. A transition matrix is row-normalised (rows sum to 1).

**Signature**
```{r}
args(simulate_matrix)
```

**Example**
```{r}
mat <- simulate_matrix(n_nodes = 4, matrix_type = "transition", seed = 1)
dim(mat)
round(mat, 3)
round(rowSums(mat), 3)                  # rows sum to 1 (empty rows -> 0)

# co-occurrence is symmetric
co <- simulate_matrix(n_nodes = 4, matrix_type = "co-occurrence", seed = 1)
isSymmetric(co)
```

## `generate_tna_matrix()`

Backward-compatibility alias of `simulate_tna_matrix()` — see that section for full documentation. It forwards all arguments unchanged.

**Signature**
```{r}
args(generate_tna_matrix)
```

**Example**
```{r}
net <- generate_tna_matrix(nodes_per_group = 2, n_groups = 2, seed = 1, verbose = FALSE)
identical(names(net), c("matrix", "node_types"))
dim(net$matrix)
```

## `generate_tna_networks()`

Backward-compatibility alias of `simulate_tna_networks()` — see that section for full documentation. It forwards all arguments unchanged.

**Signature**
```{r}
args(generate_tna_networks)
```

**Example**
```{r}
nets <- generate_tna_networks(
  n_networks = 2, n_states = 4, n_sequences = 30, seq_length = 10,
  seed = 1, verbose = FALSE
)
length(nets)
class(nets[[1]])
```

## `generate_tna_datasets()`

Backward-compatibility alias of `simulate_tna_datasets()` — see that section for full documentation. It forwards all arguments unchanged.

**Signature**
```{r}
args(generate_tna_datasets)
```

**Example**
```{r}
data <- generate_tna_datasets(
  n_datasets = 2, n_states = 4, n_sequences = 30, seq_length = 10,
  seed = 1, verbose = FALSE
)
length(data)
names(data[[1]])
```

## `generate_group_tna_networks()`

Backward-compatibility alias of `simulate_group_tna_networks()` — see that section for full documentation. It forwards all arguments unchanged.

**Signature**
```{r}
args(generate_group_tna_networks)
```

**Example**
```{r}
g <- generate_group_tna_networks(
  n_groups = 2, n_actors = 5, n_states = 3, seq_length_range = c(5, 8),
  seed = 1, verbose = FALSE
)
class(g)
length(g)
```

## `generate_probabilities()`

Generate random transition probabilities and initial-state probabilities for a Markov chain (via `seqHMM`). Returns `list(initial_probs, transition_probs, state_names)`; transition-matrix rows sum to 1.

**Signature**
```{r}
args(generate_probabilities)
```

**Example**
```{r}
probs <- generate_probabilities(n_states = 4, seed = 1)
names(probs)
probs$state_names
round(probs$initial_probs, 3)
sum(probs$initial_probs)                # initial probs sum to 1
round(probs$transition_probs, 3)
round(rowSums(probs$transition_probs), 3)
```

## `generate_sequence_data()`

Backward-compatibility alias of `simulate_tna_datasets()` — see that section for full documentation. It forwards all arguments unchanged, returning the full datasets bundle (model + probabilities + sequences).

**Signature**
```{r}
args(generate_sequence_data)
```

**Example**
```{r}
seqs <- generate_sequence_data(
  n_datasets = 1, n_states = 4, n_sequences = 30, seq_length = 10,
  seed = 1, verbose = FALSE
)
length(seqs)
names(seqs[[1]])
head(seqs[[1]]$sequences, 3)
```

## `sample_tna()`

Take a fitted `tna` model (or a sequence `data.frame`), randomly sample a proportion of its rows, and re-estimate a new `tna` model with the same parameters. `sampling_percent` must be in `(0, 1]`. Useful for sampling-stability and bootstrap-style studies.

**Signature**
```{r}
args(sample_tna)
```

**Example**
```{r}
model   <- simulate_tna_network(n_states = 4, seed = 1)
sampled <- sample_tna(model, sampling_percent = 0.5)
class(sampled)
nrow(model$data)                        # rows in the original model
nrow(sampled$data)                      # ~50% of the rows, re-estimated
head(round(sampled$weights, 3))
```
