---
title: "Saqrlab Reference — Batch, grid & sampling"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This page covers Saqrlab's orchestration layer: tools that run a simulation
many times across a parameter grid, bootstrap edge-significance recovery, fit
batches of TNA models, and summarize the resulting collections. Every example
here is kept deliberately tiny (small grids, `num_runs <= 3`, `num_cores = 1`)
so it runs in seconds; in practice you would use larger grids and parallel
cores. A shared two-state Markov model is reused throughout.

```{r shared-model}
trans <- matrix(c(.7, .3, .4, .6), 2, 2, byrow = TRUE,
                dimnames = list(c("A", "B"), c("A", "B")))
Model <- list(weights = trans, inits = c(A = 0.5, B = 0.5))
Model
```

## `run_grid_simulation()`

Runs a TNA bootstrap recovery experiment across a grid of sample sizes,
sequence lengths and missingness levels, returning one result element per
parameter combination.

**Signature**
```{r}
args(run_grid_simulation)
```

**Example**
```{r}
grid <- run_grid_simulation(
  Model = Model,
  stable_transitions = list(c("A", "B")),
  num_runs = 2,
  n_sequences_vec = 20,
  seq_length_vec = 10,
  num_cores = 1
)
length(grid)
names(grid[[1]])
```

## `summarize_grid_results()`

Aggregates a `run_grid_simulation()` result into overall-performance, edge-level
and per-setting summaries; set `print_output = FALSE` to get the data back
silently.

**Signature**
```{r}
args(summarize_grid_results)
```

**Example**
```{r}
grid_summary <- summarize_grid_results(grid, print_output = FALSE)
names(grid_summary)
grid_summary$n_selected
```

## `analyze_grid_results()`

Backward-compatible alias of `summarize_grid_results()` — identical behaviour
and arguments.

**Signature**
```{r}
args(analyze_grid_results)
```

**Example**
```{r}
identical(
  analyze_grid_results(grid, print_output = FALSE)$n_selected,
  summarize_grid_results(grid, print_output = FALSE)$n_selected
)
```

## `run_bootstrap_simulation()`

Repeats a single bootstrap edge-recovery experiment `num_runs` times for one
fixed parameter setting and aggregates how reliably the known stable edges are
recovered.

**Signature**
```{r}
args(run_bootstrap_simulation)
```

**Example**
```{r}
boot <- run_bootstrap_simulation(
  Model = Model,
  stable_transitions = list(c("A", "B")),
  num_runs = 2,
  seq_length = 10,
  n_sequences = 20,
  num_cores = 1
)
names(boot)
boot$successful_runs
```

## `run_bootstrap_iteration()`

The single-iteration engine underneath `run_bootstrap_simulation()`: simulates
one dataset from a transition matrix + initial probabilities, fits a model,
bootstraps it, and scores edge recovery (TP/TN/FP/FN) against the known stable
edges.

**Signature**
```{r}
args(run_bootstrap_iteration)
```

**Example**
```{r}
iter <- run_bootstrap_iteration(
  trans_matrix = trans,
  init_probs = c(A = 0.5, B = 0.5),
  stable_transitions = list(c("A", "B")),
  seq_length = 10,
  n_sequences = 20
)
names(iter)
iter$per_edge
```

## `evaluate_bootstrap()`

Backward-compatible alias of `run_bootstrap_iteration()`.

**Signature**
```{r}
args(evaluate_bootstrap)
```

**Example**
```{r}
ev <- evaluate_bootstrap(
  trans_matrix = trans,
  init_probs = c(A = 0.5, B = 0.5),
  stable_transitions = list(c("A", "B")),
  seq_length = 10,
  n_sequences = 20
)
names(ev)
```

## `run_network_simulation()`

Simulates many replicate datasets from observed sequence data, fits one or more
TNA model types to each, and returns tidy per-replicate `metrics` plus
aggregated `summary_stats`.

**Signature**
```{r}
args(run_network_simulation)
```

**Example**
```{r}
original_data <- simulate_sequences(
  trans_matrix = trans, init_probs = c(A = 0.5, B = 0.5),
  seq_length = 12, n_sequences = 30
)
net_sim <- run_network_simulation(
  original_data_list = original_data,
  sim_params = list(seq_length = 12, n_sequences = 30),
  models = c("tna"),
  comparisons = c("original"),
  num_runs = 3
)
names(net_sim)
head(net_sim$metrics)
```

## `summarize_simulation()`

Summarizes a tidy simulation-metrics `data.frame` (such as
`run_network_simulation()$metrics`), optionally grouped by one or more columns,
reporting mean/sd/CI per numeric column.

**Signature**
```{r}
args(summarize_simulation)
```

**Example**
```{r}
summarize_simulation(net_sim$metrics, by = "model_type")
```

## `summarize_networks()`

Summarizes a list of fitted TNA models, reporting density, centrality and
edge-count statistics per network plus an aggregate row.

**Signature**
```{r}
args(summarize_networks)
```

**Example**
```{r}
datasets <- lapply(1:2, function(i) {
  simulate_sequences(trans_matrix = trans, init_probs = c(A = 0.5, B = 0.5),
                     seq_length = 12, n_sequences = 30)
})
models <- batch_fit_models(datasets, model_type = "tna", progress = FALSE)
net_summary <- summarize_networks(models)
names(net_summary)
net_summary$n_networks
```

## `batch_fit_models()`

Fits a TNA model type to each dataset in a list, returning a list of fitted
models (with `NULL` entries where fitting failed). Supports optional parallel
execution.

**Signature**
```{r}
args(batch_fit_models)
```

**Example**
```{r}
models <- batch_fit_models(datasets, model_type = "tna", progress = FALSE)
length(models)
vapply(models, function(m) if (!is.null(m)) "OK" else "Failed", character(1))
```

## `batch_apply()`

Applies a function over a list of objects (e.g. fitted models), with optional
parallelism, progress reporting and simplification — a thin, list-friendly
wrapper for batch extraction tasks.

**Signature**
```{r}
args(batch_apply)
```

**Example**
```{r}
edge_counts <- batch_apply(
  models,
  function(m) sum(m$weights > 0, na.rm = TRUE),
  progress = FALSE
)
unlist(edge_counts)
```

## `generate_param_grid()`

Builds a parameter grid (one row per parameter combination) via random, Latin
Hypercube, or regular-grid sampling. With no arguments it returns a small demo
grid of default TNA parameters.

**Signature**
```{r}
args(generate_param_grid)
```

**Example**
```{r}
head(generate_param_grid())

ranges <- list(num_rows = c(50, 100), max_seq_length = c(10, 20),
               stability_prob = c(0.7, 1.0))
generate_param_grid(ranges, n = 4, method = "lhs")
```

## `create_param_grid()`

Backward-compatible alias of `generate_param_grid()`.

**Signature**
```{r}
args(create_param_grid)
```

**Example**
```{r}
create_param_grid(list(num_rows = c(50, 100)), n = 3, method = "random")
```

## `run_sampling_analysis()`

Repeatedly subsamples the sequence data underlying a fitted TNA model, refits on
each subsample, and reports the stability of edge weights and centralities
across iterations.

**Signature**
```{r}
args(run_sampling_analysis)
```

**Example**
```{r}
suppressMessages(library(tna))
fit <- tna(group_regulation)
samp <- run_sampling_analysis(fit, iterations = 5, sampling_percent = 0.3,
                              seed = 1, verbose = FALSE)
names(samp)
head(samp$aggregated)
```
