---
title: "Saqrlab Reference — Laboratory infrastructure"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This page documents Saqrlab's "laboratory" layer — the unified front door to the
simulators. `simulate()` dispatches to any registered simulator by name and
returns a tidy `saqr_sim` object; `validate_recovery()` checks how well an
estimator recovers known ground-truth parameters; and the scenario system
(`list_scenarios()` / `get_scenario()` / `run_scenario()`) packages named,
multi-case experiments that can be tidied and exported to disk.

## `saqr_sim()`

Constructs the standard `saqr_sim` result object that wraps a simulated
`data` frame together with its ground-truth `params`, the simulator `type`, and
the `seed`. Every explicit-parameter simulator returns one of these.

**Signature**
```{r}
args(saqr_sim)
```

**Example**
```{r}
sim <- saqr_sim(
  data = data.frame(x = rnorm(10), y = rnorm(10)),
  params = list(mean_x = 0, mean_y = 0),
  type = "demo", seed = 1
)
sim
```

The class carries a full set of S3 methods so it behaves like a tidy result:

```{r}
class(sim)
names(sim)             # data, params
dim(sim)               # dims of the underlying data frame
head(sim)              # head of $data
summary(sim)           # data summary + parameter list
sim$data |> head(3)    # the simulated data frame
sim$params             # the ground-truth parameters
as.data.frame(sim) |> head(3)  # tidy accessor -> the data frame
```

## `simulate()`

The universal dispatcher: `simulate(type, ...)` routes to the named simulator
with explicit named arguments and returns a `saqr_sim`. The `seed` argument
makes any draw reproducible.

**Signature**
```{r}
args(simulate)
```

**Example**
```{r}
ttest_sim <- simulate("ttest", n_a = 50, n_b = 50,
                      mean_a = 0, mean_b = 0.5, seed = 1)
ttest_sim
head(ttest_sim)
```

## `list_simulators()`

Returns the catalogue of registered simulator names with their descriptions as
a tidy `data.frame`.

**Signature**
```{r}
args(list_simulators)
```

**Example**
```{r}
list_simulators()
```

## `validate_recovery()`

Compares a vector of estimated parameters against the ground-truth `params`
stored in a `saqr_sim`, returning a `recovery_result` table of absolute and
relative error with a within-tolerance flag per parameter.

**Signature**
```{r}
args(validate_recovery)
```

**Example**

End-to-end: simulate a regression with known coefficients, fit it with `lm()`,
and feed the fitted coefficients back in as estimates. The estimate names must
match the `coefs.*` parameter names stored by `simulate_regression()`.

```{r}
rsim <- simulate_regression(
  coefs = c("(Intercept)" = 1, x1 = 2, x2 = -1.5),
  predictor_sds = c(x1 = 1, x2 = 1),
  error_sd = 1, n = 2000, seed = 7
)
names(rsim$params)   # confirm the coefs.* parameter names

fit_lm <- lm(y ~ x1 + x2, data = rsim$data)
est <- stats::setNames(coef(fit_lm), paste0("coefs.", names(coef(fit_lm))))

recovery <- validate_recovery(rsim, estimates = est)
recovery
summary(recovery)
```

## `list_scenarios()`

Lists the built-in named scenarios — pre-packaged, multi-case experiments — as a
tidy `data.frame`.

**Signature**
```{r}
args(list_scenarios)
```

**Example**
```{r}
list_scenarios()
```

## `get_scenario()`

Returns the parameter specification(s) for a named scenario: the full list of
cases, or a single case when `case` is supplied.

**Signature**
```{r}
args(get_scenario)
```

**Example**
```{r}
power_spec <- get_scenario("power_ttest")
length(power_spec)        # one entry per case
power_spec[[1]]           # the first case's parameters

get_scenario("power_ttest", case = 1)
```

## `run_scenario()`

Executes every case of a named scenario with a single `seed`, returning a list
of per-case `saqr_sim` results.

**Signature**
```{r}
args(run_scenario)
```

**Example**
```{r}
scenario_run <- run_scenario("power_ttest", seed = 1)
names(scenario_run)       # one element per case
scenario_run[[1]]
```

## `tidy_simulation_results()`

Flattens a `run_scenario()` result into a single tidy `data.frame`, stacking the
cases with a `.case` column and joining in each case's parameters.

**Signature**
```{r}
args(tidy_simulation_results)
```

**Example**
```{r}
tidy <- tidy_simulation_results(scenario_run)
head(tidy)
```

## `export_simulation()`

Writes a scenario run (or its tidied form) to a CSV file on disk, returning the
path invisibly.

**Signature**
```{r}
args(export_simulation)
```

**Example**
```{r}
out_path <- tempfile(fileext = ".csv")
export_simulation(tidy, out_path)
file.exists(out_path)
head(read.csv(out_path))
```
