---
title: "Saqrlab Reference — Latent-variable models"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

The latent-variable simulators generate data with known latent structure:
latent profiles (LPA) and classes (LCA), linear regression, factor models, and
sequence clusters governed by per-cluster Markov chains. Each returns a
`saqr_sim` object exposing `$data` and the explicit ground-truth `$params`, so
parameter recovery can be checked directly.

## `simulate_lpa()`

Generate continuous-indicator data with known latent profile structure. All
distributional parameters are specified explicitly so that recovery can be
verified.

**Signature**
```{r}
args(simulate_lpa)
```

**Example**
```{r}
means <- matrix(c(0, 0, 10, 10), nrow = 2, ncol = 2)
r <- simulate_lpa(means = means, sds = 0.5, props = c(0.5, 0.5), n = 200, seed = 1)
head(r$data)
str(r$params)
```

```{r lpa-plot, fig.cap="First two indicators coloured by true profile."}
lab <- r$data[[grep("class|profile|true", names(r$data), value = TRUE)[1]]]
plot(r$data[[1]], r$data[[2]], col = as.integer(factor(lab)), pch = 19,
     xlab = names(r$data)[1], ylab = names(r$data)[2],
     main = "simulate_lpa(): two latent profiles")
```

## `simulate_lca()`

Generate binary-indicator data with known latent class structure. All item
response probabilities are specified explicitly.

**Signature**
```{r}
args(simulate_lca)
```

**Example**
```{r}
item_probs <- matrix(c(0.9, 0.1, 0.9, 0.1,
                       0.1, 0.9, 0.1, 0.9), nrow = 4, ncol = 2)
r <- simulate_lca(item_probs = item_probs, class_probs = c(0.5, 0.5),
                  n = 200, seed = 1)
head(r$data)
str(r$params)
```

## `simulate_regression()`

Generate a linear-regression dataset with fully specified ground-truth
coefficients and predictor standard deviations, designed so that
`lm(y ~ ., data = r$data)` recovers the true coefficients at large `n`.

**Signature**
```{r}
args(simulate_regression)
```

**Example**
```{r}
coefs <- c("(Intercept)" = 2, x1 = 3, x2 = -1)
r <- simulate_regression(coefs = coefs,
                         predictor_sds = c(x1 = 1, x2 = 1),
                         error_sd = 0.5, n = 500, seed = 42)
head(r$data)
coef(lm(y ~ ., data = r$data))   # should be close to c(2, 3, -1)
str(r$params)
```

## `simulate_fa()`

Generate multivariate normal data from an explicit factor model. The implied
covariance matrix \(\Sigma = \Lambda \Phi \Lambda' + \Psi\) is fully specified
and stored for direct comparison with estimated covariances.

**Signature**
```{r}
args(simulate_fa)
```

**Example**
```{r}
# Two orthogonal factors, 6 indicators (3 per factor)
loadings <- matrix(c(0.8, 0.7, 0.6, 0,   0,   0,
                     0,   0,   0,   0.8, 0.7, 0.6),
                   nrow = 6, ncol = 2)  # byrow = FALSE (default)
r <- simulate_fa(loadings = loadings, n = 500, seed = 1)
head(r$data)                  # y1..y6
r$params$sigma_implied        # implied covariance matrix
```

## `simulate_seq_clusters()`

Generate wide-format sequence data where each row is drawn from one of K Markov
chains (clusters), each governed by its own transition matrix. Matrices can be
supplied explicitly or generated at random.

**Signature**
```{r}
args(simulate_seq_clusters)
```

**Example**
```{r}
m1 <- matrix(c(0.8, 0.2, 0.3, 0.7), nrow = 2, byrow = TRUE,
             dimnames = list(c("A","B"), c("A","B")))
m2 <- matrix(c(0.2, 0.8, 0.7, 0.3), nrow = 2, byrow = TRUE,
             dimnames = list(c("A","B"), c("A","B")))
r <- simulate_seq_clusters(trans_list = list(m1, m2), n = 100, seed = 1)
head(r$data)
str(r$params)
```
