---
title: "Saqrlab Reference — Longitudinal & multilevel"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

The longitudinal and multilevel simulators generate nested and time-structured
data with known generating parameters: VAR(1) multilevel time series for mlVAR /
RI-CLPM, two-level (cluster-nested) data with a target intraclass correlation,
and linear latent growth-curve trajectories. Each returns a `saqr_sim` object
with `$data` and the true `$params`.

## `simulate_longitudinal()`

Generate multilevel time-series data (a VAR(1) process with temporal,
contemporaneous, and between-person layers) suitable for testing mlVAR,
RI-CLPM, latent growth, and similar longitudinal models.

**Signature**
```{r}
args(simulate_longitudinal)
```

**Example**
```{r}
# 30 subjects, 60 time points, 3 variables
r <- simulate_longitudinal(n = 30, tp = 60, vars = 3, seed = 42)
head(r$data)
r$params$temporal   # the true VAR(1) matrix

# Explicit temporal structure
B <- matrix(c(0.4,  0.0, 0.1,
              0.2,  0.3, 0.0,
              0.0, -0.1, 0.5), nrow = 3, byrow = TRUE)
r2 <- simulate_longitudinal(n = 50, tp = 100, vars = 3, temporal = B, seed = 1)
str(r2$params, max.level = 1)
```

## `simulate_mlm()`

Generate two-level data where level-1 units are nested within clusters, with a
target intraclass correlation and a fixed slope on the first predictor (and an
optional cluster random slope), designed so both are recovered at large
`n_clusters`. The core example below does not require `lme4`; an optional fit is
shown afterwards.

**Signature**
```{r}
args(simulate_mlm)
```

**Example**
```{r}
r <- simulate_mlm(n_clusters = 40, cluster_size = 25, slope = 0.8,
                  icc = 0.2, seed = 1)
print(r)
head(r$data)
str(r$params)
```

```{r mlm-lme4, eval=requireNamespace("lme4", quietly=TRUE)}
# Optional: recover the fixed slope and ICC with lme4
m <- lme4::lmer(y ~ x1 + (1 | cluster_id), data = r$data)
lme4::fixef(m)
print(lme4::VarCorr(m))
```

## `simulate_growth()`

Generate longitudinal data from a linear latent growth-curve model: each
subject draws a random intercept and slope from a bivariate normal, then
\(y_{it} = b_{0i} + b_{1i}\,t + e_{it}\) with `time` coded `0..(n_time - 1)`.
Returned in long format.

**Signature**
```{r}
args(simulate_growth)
```

**Example**
```{r}
r <- simulate_growth(n = 300, n_time = 6, slope_mean = 2,
                     intercept_sd = 1.5, seed = 1)
print(r)
head(r$data)
str(r$params)
```

```{r growth-plot, fig.cap="Trajectories of a few subjects."}
d <- r$data
ids <- head(unique(d$subject), 8)
sub <- d[d$subject %in% ids, ]
plot(sub$time, sub$y, type = "n",
     xlab = "time", ylab = "y",
     main = "simulate_growth(): individual trajectories")
invisible(lapply(split(sub, sub$subject), function(s) {
  s <- s[order(s$time), ]
  lines(s$time, s$y, col = as.integer(factor(s$subject[1], levels = ids)), lwd = 2)
}))
```
