---
title: "Saqrlab Reference — Survival & hidden Markov"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers two time-to-event / latent-process simulators.
`simulate_survival()` draws right-censored survival data from a Cox
proportional-hazards generative model with known log hazard ratios, and
`simulate_hmm()` produces observed symbol sequences from a discrete hidden
Markov model with known transition and emission matrices. Both return a
`saqr_sim` object whose `$params` hold the ground truth.

## `simulate_survival()`

Generate right-censored survival data from a Cox proportional-hazards model
with a parametric baseline hazard (Weibull, exponential, or Gompertz). The data
are built so that `coxph(Surv(time, status) ~ ., data)` recovers the true
`betas` at large `n`.

**Signature**
```{r}
args(simulate_survival)
```

**Example**

The data frame has `time`, a `status` indicator (1 = event, 0 = censored), and
the covariates `X1 ... Xp`.
```{r}
r <- simulate_survival(n = 400, n_covariates = 2, seed = 1)
r
head(as.data.frame(r))
r$params$betas
r$params$realized_censoring_rate
```

A second configuration with binary covariates and an exponential baseline:
```{r}
r2 <- simulate_survival(n = 500, betas = c(0.8, -0.5),
                        baseline = "exponential",
                        covariate_type = "binary", seed = 42)
summary(r2)
```

**Visualising event vs. censoring times**

Overlaid histograms of observed times, split by whether the observation was an
event or was censored, show the censoring structure.
```{r}
surv_df <- as.data.frame(r)
ev   <- surv_df$time[surv_df$status == 1]
cens <- surv_df$time[surv_df$status == 0]
brk  <- seq(0, max(surv_df$time) + 1, length.out = 25)

hist(ev, breaks = brk, col = "#2c7fb880", border = "white",
     xlab = "Observed time", main = "Event vs. censoring times",
     ylim = c(0, max(hist(ev, breaks = brk, plot = FALSE)$counts,
                     hist(cens, breaks = brk, plot = FALSE)$counts)))
hist(cens, breaks = brk, col = "#d95f0e80", border = "white", add = TRUE)
legend("topright", fill = c("#2c7fb880", "#d95f0e80"),
       legend = c("event (status = 1)", "censored (status = 0)"), bty = "n")
```

**Optional: Kaplan–Meier and recovery with `survival`**

When the `survival` package is installed, a Kaplan–Meier curve and a Cox fit
confirm the generating effects. Skipped if `survival` is unavailable.
```{r km-cox, eval=requireNamespace("survival", quietly=TRUE)}
km <- survival::survfit(survival::Surv(time, status) ~ 1, data = surv_df)
plot(km, col = "#2c7fb8", lwd = 2, xlab = "Time", ylab = "Survival probability",
     main = "Kaplan–Meier estimate")

cox <- survival::coxph(survival::Surv(time, status) ~ X1 + X2, data = surv_df)
data.frame(true = r$params$betas, estimated = unname(coef(cox)))
```

## `simulate_hmm()`

Generate observed symbol sequences from a discrete hidden Markov model. For each
sequence a latent state path is drawn from the Markov chain (`init`, `trans`),
and each hidden state emits an observed symbol via the `emission` matrix. The
true latent paths are returned in `$params$hidden_paths`.

**Signature**
```{r}
args(simulate_hmm)
```

**Example**

The data are long-format: one row per (`sequence_id`, `time`) with the observed
`symbol`. The true transition and emission matrices live in `$params`.
```{r}
h <- simulate_hmm(n_sequences = 100, seq_length = 40, seed = 1)
h
head(as.data.frame(h))
h$params$trans
h$params$emission
```

A second, explicit three-state model:
```{r}
tr <- matrix(c(0.8, 0.1, 0.1,
               0.1, 0.8, 0.1,
               0.1, 0.1, 0.8), nrow = 3, byrow = TRUE)
em <- matrix(c(0.7, 0.2, 0.1,
               0.1, 0.7, 0.2,
               0.2, 0.1, 0.7), nrow = 3, byrow = TRUE)
h2 <- simulate_hmm(n_sequences = 80, seq_length = 50, n_states = 3,
                   n_symbols = 3, trans = tr, emission = em, seed = 42)
summary(h2)
dim(h2$params$hidden_paths)
h2$params$hidden_paths[1:3, 1:10]
```
