---
title: "Saqrlab Reference — Visualization"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
suppressMessages(devtools::load_all("/Users/mohammedsaqr/Documents/Github/Saqrlab", quiet = TRUE))
set.seed(1)
```

This reference covers Saqrlab's `ggplot2`-based plotting verbs for comparison
results. `plot_network_estimation()` visualizes the output of
`compare_network_estimation()`, `plot_tna_comparison()` plots the per-iteration
metric distributions from a comparison, and `plot_sampling_distribution()` draws
a single metric's distribution from `run_sampling_analysis()`. Each example first
builds the minimal result object the plot consumes, then renders the figure.

```{r build-inputs}
# Shared inputs: simulate sequences, fit a model, and run the comparisons
# that the plotting verbs consume.
set.seed(1)
trans_mat <- matrix(c(
  0.7, 0.2, 0.1,
  0.3, 0.5, 0.2,
  0.2, 0.3, 0.5
), nrow = 3, byrow = TRUE)
rownames(trans_mat) <- colnames(trans_mat) <- c("A", "B", "C")
init_probs <- c(A = 0.5, B = 0.3, C = 0.2)

seqs <- simulate_sequences(
  transition_matrix = trans_mat,
  initial_probabilities = init_probs,
  max_seq_length = 12,
  num_rows = 40
)

estimation <- compare_network_estimation(
  seqs,
  model_types = c("tna", "ftna"),
  iterations = 15,
  seed = 1,
  verbose = FALSE
)

model <- fit_network_model(seqs, "tna")
sampling <- run_sampling_analysis(model, iterations = 15, seed = 1, verbose = FALSE)
```

## `plot_network_estimation()`

Plot the distribution of a metric (default Pearson) across sampling iterations for
each model type from a `compare_network_estimation()` result, returning a `ggplot`
object.

**Signature**
```{r}
args(plot_network_estimation)
```

**Example**
```{r plot-network-estimation, eval=requireNamespace("ggplot2", quietly = TRUE)}
plot_network_estimation(estimation, metric_name = "Pearson", plot_type = "histogram")
```

## `plot_sampling_distribution()`

Plot the distribution of a single metric within a category (e.g. Pearson within
"Correlations") from the `individual` data frame of a `run_sampling_analysis()`
result, returning a `ggplot` object.

**Signature**
```{r}
args(plot_sampling_distribution)
```

**Example**
```{r plot-sampling-distribution, eval=requireNamespace("ggplot2", quietly = TRUE)}
plot_sampling_distribution(
  sampling$individual,
  category = "Correlations",
  metric = "Pearson"
)
```

## `plot_tna_comparison()`

Plot per-iteration comparison metrics (histogram, boxplot, ridgeline, bar, or
density) from the `individual` data frame of a comparison result, returning a
`ggplot` object.

**Signature**
```{r}
args(plot_tna_comparison)
```

**Example**
```{r plot-tna-comparison, eval=requireNamespace("ggplot2", quietly = TRUE)}
plot_tna_comparison(estimation$individual, plot_type = "boxplot", metric_name = "Pearson")
```
