---
title: "mlVAR Network Analysis — Soyster et al. (2022)"
subtitle: "Alcohol Use & Affect: Temporal, Contemporaneous, and Between-Subjects Networks"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 9, dpi = 150
)
devtools::load_all("..", quiet = TRUE)
```

## Data

**openESM 0035** (Soyster et al., 2022): 33 participants, ~129 ESM measurements
each. Six affect/craving variables measured repeatedly throughout the day.

```{r data}
ds <- openesm::get_dataset("0035_soyster")
d <- as.data.frame(ds$data)
d <- d[!is.na(d$day), ]
d <- d[order(d$id, d$day, d$creation_time), ]
d$beep <- as.integer(ave(seq_len(nrow(d)), d$id, d$day, FUN = seq_along))

vars <- c("stressed", "depressed", "calm", "happy", "angry", "craving")

cat(sprintf("Participants: %d\n", length(unique(d$id))))
cat(sprintf("Observations: %d\n", nrow(d)))
cat(sprintf("Variables: %s\n", paste(vars, collapse = ", ")))
```

## Estimation

```{r fit}
fit <- mlvar(d, vars = vars, id = "id", day = "day", beep = "beep")
print(fit)
```

```{r shared-setup}
# --- Shared visual configuration ---

# Short labels for cleaner display
short_labels <- c("STR", "DEP", "CAL", "HAP", "ANG", "CRV")

# Node colors by variable type:
#   Negative affect (stressed, depressed, angry) = warm tones
#   Positive affect (calm, happy)                = cool tones
#   Craving                                      = amber
node_colors <- c(
  "#E74C3C",   # stressed  — red

"#8E44AD",   # depressed — purple
  "#2ECC71",   # calm      — green
  "#3498DB",   # happy     — blue
  "#E67E22",   # angry     — orange
  "#F1C40F"    # craving   — gold
)

# Consistent circular layout so node positions are comparable across networks
n_vars <- length(vars)
angles <- seq(0, 2 * pi, length.out = n_vars + 1)[seq_len(n_vars)]
# Rotate so first node is at top
angles <- angles - pi / 2
shared_layout <- cbind(cos(angles), sin(angles))

# Threshold temporal network to significant edges (p < 0.05)
B <- fit$temporal
sig_mask <- matrix(TRUE, nrow(B), ncol(B))
for (k in seq_along(vars)) {
  sig_mask[k, ] <- fit$coefs[[k]]$p < 0.05
}
B_sig <- B * sig_mask
rownames(B_sig) <- short_labels
colnames(B_sig) <- short_labels

# Label the other matrices too
contemp <- fit$contemporaneous
rownames(contemp) <- short_labels
colnames(contemp) <- short_labels

between <- fit$between
rownames(between) <- short_labels
colnames(between) <- short_labels
```

## Temporal Network

Directed lag-1 regression coefficients. An edge from node *j* to node *i* means
that variable *j* at time *t*−1 predicts variable *i* at time *t*. Only
statistically significant edges shown (*p* < .05).

```{r temporal, fig.width=9, fig.height=9}
cograph::splot(
  B_sig,
  layout = shared_layout,
  directed = TRUE,
  title = "Temporal Network (lag-1, p < .05)",

  # Nodes
  node_size = 8,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1.5,
  label_size = 0.9,
  label_color = "white",
  label_fontface = "bold",

  # Edges — teal positive, crimson negative
  edge_positive_color = "#1ABC9C",
  edge_negative_color = "#E74C3C",
  edge_width_range = c(0.5, 5),
  edge_alpha = 0.85,
  arrow_size = 1.2,

  # Edge labels
  edge_labels = TRUE,
  edge_label_size = 0.75,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  edge_label_fontface = "bold",

  # Background
  background = "#FAFAFA"
)
```

## Contemporaneous Network

Undirected partial correlations among within-person residuals, estimated via
EBIC-GLASSO. These capture same-time-point associations after removing
temporal (lagged) effects.

```{r contemporaneous, fig.width=9, fig.height=9}
cograph::splot(
  contemp,
  layout = shared_layout,
  directed = FALSE,
  title = "Contemporaneous Network (EBIC-GLASSO)",

  # Nodes
  node_size = 8,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1.5,
  label_size = 0.9,
  label_color = "white",
  label_fontface = "bold",

  # Edges — blue positive, red negative
  edge_positive_color = "#2980B9",
  edge_negative_color = "#C0392B",
  edge_width_range = c(0.5, 6),
  edge_alpha = 0.8,

  # Edge labels
  edge_labels = TRUE,
  edge_label_size = 0.75,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  edge_label_fontface = "bold",

  # Background
  background = "#FAFAFA"
)
```

## Between-Subjects Network

Undirected partial correlations among person-level means, estimated via
EBIC-GLASSO. These represent stable trait-level associations across
individuals.

```{r between, fig.width=9, fig.height=9}
cograph::splot(
  between,
  layout = shared_layout,
  directed = FALSE,
  title = "Between-Subjects Network (EBIC-GLASSO)",

  # Nodes
  node_size = 8,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1.5,
  label_size = 0.9,
  label_color = "white",
  label_fontface = "bold",

  # Edges — green positive, rose negative
  edge_positive_color = "#27AE60",
  edge_negative_color = "#E74C3C",
  edge_width_range = c(0.5, 6),
  edge_alpha = 0.8,

  # Edge labels
  edge_labels = TRUE,
  edge_label_size = 0.75,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  edge_label_fontface = "bold",

  # Background
  background = "#FAFAFA"
)
```

## All Three Networks

```{r triptych, fig.width=18, fig.height=7}
old_par <- par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))

# --- Temporal ---
cograph::splot(
  B_sig,
  layout = shared_layout,
  directed = TRUE,
  title = "Temporal",
  node_size = 7,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#1ABC9C",
  edge_negative_color = "#E74C3C",
  edge_width_range = c(0.4, 4),
  edge_alpha = 0.85,
  arrow_size = 1,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  background = "#FAFAFA"
)

# --- Contemporaneous ---
cograph::splot(
  contemp,
  layout = shared_layout,
  directed = FALSE,
  title = "Contemporaneous",
  node_size = 7,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#2980B9",
  edge_negative_color = "#C0392B",
  edge_width_range = c(0.4, 5),
  edge_alpha = 0.8,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  background = "#FAFAFA"
)

# --- Between ---
cograph::splot(
  between,
  layout = shared_layout,
  directed = FALSE,
  title = "Between-Subjects",
  node_size = 7,
  node_fill = node_colors,
  node_border_color = "gray20",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#27AE60",
  edge_negative_color = "#E74C3C",
  edge_width_range = c(0.4, 5),
  edge_alpha = 0.8,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "gray20",
  edge_label_bg = "white",
  background = "#FAFAFA"
)

par(old_par)
```

## Dark Theme Variant

```{r dark-triptych, fig.width=18, fig.height=7}
dark_bg <- "#1C1C2E"
dark_node_colors <- c(
  "#FF6B6B",   # stressed
  "#C084FC",   # depressed
  "#4ADE80",   # calm
  "#60A5FA",   # happy
  "#FB923C",   # angry
  "#FACC15"    # craving
)

old_par <- par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))

cograph::splot(
  B_sig,
  layout = shared_layout,
  directed = TRUE,
  title = "Temporal",
  title_size = 1.4,
  node_size = 7,
  node_fill = dark_node_colors,
  node_border_color = "#FFFFFF40",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#34D399",
  edge_negative_color = "#FB7185",
  edge_width_range = c(0.5, 4),
  edge_alpha = 0.9,
  arrow_size = 1,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "#E2E8F0",
  edge_label_bg = "#2D2D44",
  background = dark_bg
)

cograph::splot(
  contemp,
  layout = shared_layout,
  directed = FALSE,
  title = "Contemporaneous",
  title_size = 1.4,
  node_size = 7,
  node_fill = dark_node_colors,
  node_border_color = "#FFFFFF40",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#60A5FA",
  edge_negative_color = "#FB7185",
  edge_width_range = c(0.5, 5),
  edge_alpha = 0.9,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "#E2E8F0",
  edge_label_bg = "#2D2D44",
  background = dark_bg
)

cograph::splot(
  between,
  layout = shared_layout,
  directed = FALSE,
  title = "Between-Subjects",
  title_size = 1.4,
  node_size = 7,
  node_fill = dark_node_colors,
  node_border_color = "#FFFFFF40",
  node_border_width = 1,
  label_size = 0.8,
  label_color = "white",
  label_fontface = "bold",
  edge_positive_color = "#4ADE80",
  edge_negative_color = "#FB7185",
  edge_width_range = c(0.5, 5),
  edge_alpha = 0.9,
  edge_labels = TRUE,
  edge_label_size = 0.6,
  edge_label_color = "#E2E8F0",
  edge_label_bg = "#2D2D44",
  background = dark_bg
)

par(old_par)
```

## Temporal Coefficients

```{r coefficients}
summary(fit)
```

## Speed

```{r speed}
t_our <- system.time(
  mlvar(d, vars = vars, id = "id", day = "day", beep = "beep")
)["elapsed"]

t_ref <- system.time(
  suppressWarnings(mlVAR::mlVAR(
    d, vars = vars, idvar = "id", dayvar = "day", beepvar = "beep",
    lags = 1, estimator = "lmer", temporal = "fixed",
    contemporaneous = "fixed", scale = FALSE, verbose = FALSE
  ))
)["elapsed"]

cat(sprintf("Saqrlab mlvar():  %.3f seconds\n", t_our))
cat(sprintf("mlVAR::mlVAR():   %.3f seconds\n", t_ref))
cat(sprintf("Speedup:          %.0fx\n", t_ref / t_our))
```
